Linear and Differential Cryptanalysis of DES

### 2009-02-02 trunk/talk/draft1.tex
Slide fixes + more conclusion

\vspace{5 mm}
\pause Define 
\begin{itemize}[<+->]
\item{$P_i$ where $i = 1, 2, \dots 16$ as the $i$-th plaintext bit}
\item{$C_i$ where $i = 1, 2, \dots 16$ as the $i$-th ciphertext bit}
\item{$U_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th input bit to the $j$-th round of SPN}
\item{$V_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th output bit of the $j$-th round of SPN}
\end{itemize}
\end{frame}

@@

\vspace{5mm}
\pause Assume the following probability distribution
\[ Pr(X_1 = i)  = \left\{ \begin{array}{ll}
         p_1 & \mbox{$i = 0$};\\
         (1 - p_1) & \mbox{$i = 1$}.\end{array} \right. \] 
\[ Pr(X_2 = i)  = \left\{ \begin{array}{ll}
         p_2 & \mbox{$i = 0$};\\
         (1 - p_2) & \mbox{$i = 1$}.\end{array} \right. \] 
\end{frame}

@@

\begin{frame}
Using this principle we can write the following equation over 3 rounds of SPN:
\[ U_{4,6} \oplus U_{4,8} \oplus U_{4,14} \oplus U_{4, 16} \oplus P_5 \oplus P_7 \oplus P_8 = \Sigma K \]
\pause Where $\Sigma K$ is the sum over some key bits. Note that Since the key is fixed, $\Sigma K = 0$ or $1$ and thus we can ignore it since we only care about the bias.\\
\vspace{5mm}
\pause The magnitude of the bias of the above expression, by the Piling Up Principle, is $1/32$. \\
\vspace{5mm}
\pause Next we show how we can extract key bits using this information.
\end{frame}

@@

\vspace{5mm}
\pause Using the S-Box 5 approximation and the Piling Up Principle we can combine the approximations to get
\begin{align*}
 P_L[25] & \oplus P_L[14] \oplus P_L[8] \oplus P_L[3] \oplus \\
         & U_{4L}[25] \oplus U_{4L}[14] \oplus U_{4L}[8] \oplus U_{4L}[3] \oplus \\
			& P_R[17] \oplus U_{4R}[17] = K_1[26] \oplus K_3[26] \\
\end{align*}

@@

\begin{frame}
We perform the attack on the 4-round DES similarly to how we performed the attack on 4-round SPN, by ``undoing'' the last round.\\
\vspace{5mm}
\pause
Note that in the last round, $U_{4L}$ bits are not changed, only $U_{4R}$ bits are affected. Note also that there is only one $U_{4R}$ bit the 3-round equation.
\begin{align*} 
P_L[25] \oplus P_L[14] \oplus P_L[8] \oplus & P_L[3] \oplus U_{4L}[25] \oplus U_{4L}[14] \oplus U_{4L}[8] \\ & \oplus U_{4L}[3] \oplus P_R[17] \oplus U_{4R}[17] = K_1[26] \oplus K_3[26]
\end{align*}
\vspace{5mm}
\pause
This bit, $U_{4R}[17]$, is affected by the result of S-Box 1 in round 4. If we guess correctly for the 6 partial subkey bits $K_{4,1}$ to $K_{4,6}$, the 3-round equation will hold with high bias, otherwise it will likely hold with close to 0 bias.\\
\vspace{5mm}
\pause
Thus, this attack will give us $6$ of the $56$ key bits. \pause What this means in terms of time saved over exhaustive search is this: If it takes a year to find a key through exhaustive search of $2^{56}$ bits, after this attack it will only take $5$ days to exhaustively search the remaining $2^{50}$ bits.
\end{frame}

@@

\begin{frame}
What if that is false ? ... Discussion ...

\subsection{Links between the Attacks}
\begin{frame}
Linear Resistance:\\
\vspace{5mm}
Thm: A function $F: GF(2)^p \rightarrow GF(2)^q$is perfect nonlinear (ie bias is 0) if and only if it is \textit{bent}. \\
\vspace{3mm}
Thm: Bent functions exist only for $p \geq 2q$ and $p$ even. 
\vspace{5mm}
\end{frame}

\begin{frame}
Mathematically and while implementing ...

@@

\begin{frame}
References:\\
\vspace{5mm}
Matsui\\
Beham, Shamir\\
Heys
Chabaud, Vaudenay\\
Manferdelli
\end{frame}

>>>>

\vspace{5 mm}
\pause Define 
\begin{itemize}[<+->]
\item{$P_i$ where $i = 1, 2, \dots 16$ as the $i$-th plaintext bit}
\item{$C_i$ where $i = 1, 2, \dots 16$ as the $i$-th ciphertext bit}
\item{$U_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th input bit to the $j$-th round of SPN}
\item{$V_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th output bit of the $j$-th round of SPN}
\item{$K_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th bit of the round key for the $j$-th round of SPN}
\end{itemize}
\end{frame}

@@

\vspace{5mm}
\pause Assume the following probability distribution
\[ Pr(X_1 = i)  = \left\{ \begin{array}{ll}
         p_1 & \mbox{$i = 0$}\\
         (1 - p_1) & \mbox{$i = 1$}.\end{array} \right. \] 
\[ Pr(X_2 = i)  = \left\{ \begin{array}{ll}
         p_2 & \mbox{$i = 0$};\\
         (1 - p_2) & \mbox{$i = 1$}.\end{array} \right. \] 
\end{frame}

@@

\begin{frame}
Using this principle we can write the following equation over 3 rounds of SPN:
\[ U_{4,6} \oplus U_{4,8} \oplus U_{4,14} \oplus U_{4, 16} \oplus P_5 \oplus P_7 \oplus P_8 = \Sigma K \]
\pause Where $\Sigma K$ is the sum over some key bits. Note that since the key is fixed, $\Sigma K = 0$ or $1$ and thus we can ignore it since we only care about the bias.\\
\vspace{5mm}
\pause The magnitude of the bias of the above expression, by the Piling Up Principle, is $1/32$. \\
\vspace{5mm}
\pause Next we show how we can extract key bits using this information.
\end{frame}

@@

\vspace{5mm}
\pause Using the S-Box 5 approximation and the Piling Up Principle we can combine the approximations to get
\begin{align*}
 P_L[25] & \oplus P_L[14] \oplus P_L[8] \oplus P_L[3] \oplus \\
         & U_{4R}[25] \oplus U_{4R}[14] \oplus U_{4R}[8] \oplus U_{4R}[3] \oplus \\
			& P_R[17] \oplus U_{4L}[17] = K_1[26] \oplus K_3[26] \\
\end{align*}

@@

\begin{frame}
We perform the attack on the 4-round DES similarly to how we performed the attack on 4-round SPN, by ``undoing'' the last round.\\
\vspace{5mm}
\pause
Note that in the last round, $U_{4R}$ bits are not changed, only $U_{4L}$ bits are affected. Note also that there is only one $U_{4L}$ bit the 3-round equation.
\begin{align*} 
P_L[25] \oplus P_L[14] \oplus P_L[8] \oplus & P_L[3] \oplus U_{4R}[25] \oplus U_{4R}[14] \oplus U_{4R}[8] \\ & \oplus U_{4R}[3] \oplus P_R[17] \oplus U_{4L}[17] = K_1[26] \oplus K_3[26]
\end{align*}
\vspace{5mm}
\pause
This bit, $U_{4L}[17]$, is affected by the result of S-Box 1 in round 4. If we guess correctly for the 6 partial subkey bits $K_{4,1}$ to $K_{4,6}$, the 3-round equation will hold with high bias, otherwise it will likely hold with close to 0 bias.\\
\vspace{5mm}
\pause
Thus, this attack will give us $6$ of the $56$ key bits. \pause What this means in terms of time saved over exhaustive search is this: If it takes a year to find a key through exhaustive search of $2^{56}$ bits, after this attack it will only take $5$ days to exhaustively search the remaining $2^{50}$ bits.
\end{frame}

@@

\subsection{Links between the Attacks}
\begin{frame}
Linear Resistance:\\
\vspace{3mm}
Let $F: GF(2)^p \rightarrow GF(2)^q$.\\
\vspace{3mm}
Defn: A function F is bent if its Walsh transform $\mathcal{W}\{F\}$ is $\pm 2^{p/2}$.\\
\vspace{3mm}
Thm: A function F is perfect nonlinear (ie bias is 0) if and only if it is \textit{bent}. \\
\vspace{3mm}
Thm: Bent functions exist only for $p \geq 2q$ and $p$ even. \\
\vspace{5mm}
To sumarize: When $p\geq 2q$ and $p$ even, differential-resistant is equivalent to linear resistant, and to vectorial Bentness.
\end{frame}

@@

\begin{frame}
References:\\
\vspace{5mm}
M. Matsui, "Linear Cryptanalysis Method for DES Cipher", \textit{Advances in Cryptology - EUROCRYPT '93 (Lecture Notes in Computer Science no. 765),} Springer-Verlag, pp. 38\\
\vspace{2mm}
E Biham And A. Shamir, "Differential Cryptanalysis of DES-like Cryptosystems", \textit{Journal of Cryptology}, vol 4, no 1, pp. 3-72, 1991.\\
\vspace{2mm}
H. Heys, "A Tutorial on Linear and Differential Cryptanalysis"\\
\vspace{2mm}
F. Chabaud And S. Vaudenay, "Links Between Differential and Linear Cryptanalysis", \textit{Advances in Cryptology - EUROCRYPT '94 (Lexture Notes in Computer Science no. 950)}, Springer-Verlag, pp. 356-365, 1995.\\
\vspace{2mm}
J. Manferdelli, UW Cryptography Lecture Notes
\end{frame}
